Reproducible development environments with rix

### 2024-02-01 paper.qmd
Add files via upload

There are tools available that authors can use to reach the gold stardard of a
fully reproducible study. Let us first consider the task of listing the software
used. [R]{.proglang} provides the `sessionInfo()` function whose output can be
saved into a file. Below is an example output of `sessionInfo()`:

@@

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future) can read this file and see
which version of [R]{.proglang} was used, and which packages (and their
versions) were used as well. However, others would still need to install the
correct versions of each package themselves. An alternative to this is instead
to use the [renv]{.pkg} package which generates a so-called `renv.lock` file
which also lists [R]{.proglang} and package versions. Here is an example of such
an `renv.lock` file:

@@

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. It is possible to then
execute the code of a study inside a so-called container, which is an instance
of an image. Executing, or running, these containers is done non-interactively,
and so it is possible to instantiate the correct environment to run a study in a
single command.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, as these external libraries are thus also made available to
future reproducers of the study within the [Docker]{.proglang} image. The image
can also be shared.

@@

An optional step towards the gold standard is to use a build automation tool
such as [Make]{.proglang} to run the whole analysis when the container is
executed, instead of executing scripts. Build automation tools make it easier to
run arbitrary code in a series of well-defined steps. [R]{.proglang} programmers
can use the [targets]{.pkg} by @landau2021 as a build automation tool.

>>>>

But even assuming that authors provided clear instructions to install the exact
requirements of a study, it might happen that these instructions do not even
work on different platforms, or on different versions of the same platform. And
reproducibility is not the only activity that might get impacted by a lack of
clear instructions: if an author works on several machines, the exact same
computational environment should be used on all the machines, and the same is
true when a team of authors collaborate.

There are tools available that authors can use to reach the gold stardard of a
fully reproducible study, or to successfully deploy the same development
environment on several machines. Let us first consider the task of listing the
software used. [R]{.proglang} provides the `sessionInfo()` function whose output
can be saved into a file. Below is an example output of `sessionInfo()`:

@@

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future), or co-authors, can read
this file and see which version of [R]{.proglang} was used, and which packages
(and their versions) were used as well. However, others would still need to
install the correct versions of each package manually. This can be in certain
cases challenging, especially if packages that depend on system-level
dependencies are required. For example, the [nloptr]{.pkg} package requires
either [cmake]{.proglang} on Linux or macOS, or a binary of [nlopt]{.proglang},
but this is not the case of Windows.

A better alternative to simply listing packages and their versions is to use the
[renv]{.pkg} package which generates a so-called `renv.lock` file which also
lists [R]{.proglang} and package versions. Here is an example of such an
`renv.lock` file:

@@

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. [renv]{.pkg} will also not install system-level dependencies such
as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
be handled separately, as described in [renv]{.pkg}’s Readme: 
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

Before continuing, it should be noted that other packages exist which provide
similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg} by
@simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. These steps are written
down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
definition of the [Docker]{.prolgang} image.

It is possible to then execute the code of a study inside a so-called container,
which is an instance of an image. Executing, or running, these containers is
done non-interactively, and so it is possible to instantiate the correct
environment to run a study in a single command.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete Linux
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, such as for [nloptr]{.pkg}, as these external libraries are
thus also made available to future reproducers of the study within the
[Docker]{.proglang} image. The image can also be shared.

@@

While [Docker]{.proglang} offers a great solution to the reproducibility
problem, it can be awkward to work interactively from a running container:
containers can be changed at run-time, but any change will be lost once the
container is stopped. Also, while it is possible to launch graphical programs
from a container, this is generally not easy to set up and only works, as far as
we know, only on Linux operating systems. In most cases, users of
[Docker]{.proglang} will instead use the container to run a web server and the
web based application of a popular integrated development environment (IDE). So
for instance, it would be possible to launch the web interface of the RStudio
IDE, but not the operating system client of RStudio, which is usually what an R
programmer would use (or VS Code or Emacs...). A common approach seems to be to
work on a study interactively using a standard installation of R and R packages,
and after the study is done, take some time to write the required `Dockerfile`
to provide an easy way for future replicators to reproduce the results of the
study. But this approach does not solve the problem of deploying the same
environment on several machines, and so ideally authors would develop using
[Docker]{.proglang}. Another challenge of using [Docker]{.proglang} is that,
as stated already, [Docker]{.proglang} images are actually almost complete
Linux operating systems, so experience with Linux is highly recommended if
one wishes to properly write a `Dockerfile`.

A final, and optional, step towards the gold standard is to use a build
automation tool such as [Make]{.proglang} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.
